cardinalR: Generating interesting high-dimensional data structures

### 2025-08-22 paper-cardinalR.Rmd
Updaded grided sphere section

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ observations evenly distributed on the surface of a $3\text{-}D$ unit sphere, with optional Gaussian noise dimensions when $p > 3$.

The base structure consists of a $3\text{-}D$ spherical surface created using a regular grid in spherical coordinates, where $\theta \in [0, 2\pi]$ represents the azimuthal angle and $\phi \in [0, \pi]$ denotes the polar angle.

The number of grid steps along each dimension is determined by factoring $n$ into two approximately equal integers via `gen_nproduct(n, p = 2)`.

Each point on the sphere is computed using the spherical-to-Cartesian transformation: $X_1 = \sin(\phi)\cos(\theta)$, $X_2 = \sin(\phi)\sin(\theta)$, $X_3 = \cos(\phi)$. For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

>>>>

The `gen_gridedsphere(n, p)` function generates a $p$-dimensional dataset of approximately $n$ points evenly distributed on the surface of a $(p!-!1)$-sphere (the unit sphere embedded in $\mathbb{R}^p$).

The construction is based on a regular grid in spherical coordinates with $(p-1)$ angular variables. Specifically:

* For dimensions $j = 1, \dots, p-2$, the polar angles are defined on $[0, \pi]$.
* For the last angle ($j = p-1$), the azimuthal angle is defined on $[0, 2\pi]$.

The number of grid steps along each angular dimension is determined by factoring $n$ into $(p-1)$ approximately equal integers via `gen_nproduct(n, p - 1)`.

Each grid point is then mapped to Cartesian coordinates using the standard hyperspherical-to-Cartesian transformation:

$$
\begin{aligned}
X_1 &= \cos(\theta_1), \\
X_2 &= \sin(\theta_1)\cos(\theta_2), \\
X_3 &= \sin(\theta_1)\sin(\theta_2)\cos(\theta_3), \\
&\;\;\vdots \\
X_{p-1} &= \sin(\theta_1)\sin(\theta_2)\cdots \sin(\theta_{p-2})\cos(\theta_{p-1}), \\
X_p &= \sin(\theta_1)\sin(\theta_2)\cdots \sin(\theta_{p-2})\sin(\theta_{p-1}).
\end{aligned}
$$

Thus, the output is a deterministic grid of points lying exactly on the unit $(p-1)$-sphere, with no added noise dimensions.

### 2025-08-22 paper-cardinalR.tex
Updaded grided sphere section

The \texttt{gen\_gridedsphere(n,\ p)} function generates a \(p\)-dimensional dataset of approximately \(n\) observations evenly distributed on the surface of a \(3\text{-}D\) unit sphere, with optional Gaussian noise dimensions when \(p > 3\).

The base structure consists of a \(3\text{-}D\) spherical surface created using a regular grid in spherical coordinates, where \(\theta \in [0, 2\pi]\) represents the azimuthal angle and \(\phi \in [0, \pi]\) denotes the polar angle.

The number of grid steps along each dimension is determined by factoring \(n\) into two approximately equal integers via \texttt{gen\_nproduct(n,\ p\ =\ 2)}.

Each point on the sphere is computed using the spherical-to-Cartesian transformation: \(X_1 = \sin(\phi)\cos(\theta)\), \(X_2 = \sin(\phi)\sin(\theta)\), \(X_3 = \cos(\phi)\). For \(p > 3\), the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: \(X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p\).

>>>>

The \texttt{gen\_gridedsphere(n,\ p)} function generates a \(p\)-dimensional dataset of approximately \(n\) points evenly distributed on the surface of a \((p!-!1)\)-sphere (the unit sphere embedded in \(\mathbb{R}^p\)).

The construction is based on a regular grid in spherical coordinates with \((p-1)\) angular variables. Specifically:

\begin{itemize}
\tightlist
\item
  For dimensions \(j = 1, \dots, p-2\), the polar angles are defined on \([0, \pi]\).
\item
  For the last angle (\(j = p-1\)), the azimuthal angle is defined on \([0, 2\pi]\).
\end{itemize}

The number of grid steps along each angular dimension is determined by factoring \(n\) into \((p-1)\) approximately equal integers via \texttt{gen\_nproduct(n,\ p\ -\ 1)}.

Each grid point is then mapped to Cartesian coordinates using the standard hyperspherical-to-Cartesian transformation:

\[
\begin{aligned}
X_1 &= \cos(\theta_1), \\
X_2 &= \sin(\theta_1)\cos(\theta_2), \\
X_3 &= \sin(\theta_1)\sin(\theta_2)\cos(\theta_3), \\
&\;\;\vdots \\
X_{p-1} &= \sin(\theta_1)\sin(\theta_2)\cdots \sin(\theta_{p-2})\cos(\theta_{p-1}), \\
X_p &= \sin(\theta_1)\sin(\theta_2)\cdots \sin(\theta_{p-2})\sin(\theta_{p-1}).
\end{aligned}
\]

Thus, the output is a deterministic grid of points lying exactly on the unit \((p-1)\)-sphere, with no added noise dimensions.
